---
title: "Lesson 2 — Logistic regression (classical and machine-learning style)"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The concept

## What problem does it solve?

**Logistic regression** predicts the **probability** of a yes/no outcome
(here: coronary artery disease, CAD) from several predictors. It is the
standard method in clinical research and the **baseline every ML model must
beat**.

## How it works

1. It adds up the predictors, each multiplied by a weight (a *coefficient*).
2. The total (the *score*) can be any number, so it is squeezed between 0 and
   1 by an S-shaped curve (the *logistic* or *sigmoid* function). The result
   is a probability.
3. The weights are chosen so the predicted probabilities fit the observed
   outcomes as closely as possible (*maximum likelihood*).

## The same idea as equations

**Step 1: the score** (a weighted sum of the predictors). $\beta_0$ is the
intercept, and $\beta_1, \beta_2, \dots$ are the coefficients that the model
learns from the data:

$$
\text{score} = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{male} + \dots + \beta_k \times x_k
$$

**Step 2: from score to probability.** The logistic function turns any score
into a probability $p$ between 0 and 1:

$$
p = P(\text{CAD} = \text{Yes}) = \frac{1}{1 + e^{-\text{score}}}
$$

**The same model written the other way round.** The score is the *log-odds*
of the outcome. That's why each coefficient is "the change in log-odds per
unit of the predictor":

$$
\log\left(\frac{p}{1 - p}\right) = \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{male} + \dots
$$

**Odds ratio (OR).** Exponentiating a coefficient gives its odds ratio:

$$
\text{OR}_j = e^{\beta_j}
$$

An OR of 2 for "male" means men have twice the *odds* of CAD as women,
**holding the other variables constant**.

The S-shaped curve in Step 2 looks like this. A score of 0 gives a
probability of 0.5, large positive scores approach 1, and large negative
scores approach 0:

```{r sigmoid-curve, fig.height=3.5, fig.width=6}
# curve() draws a mathematical function. x is the score; the formula is the logistic function.
curve(1 / (1 + exp(-x)), from = -6, to = 6,
      xlab = "Score (log-odds)", ylab = "Probability of CAD",
      lwd = 2, col = "steelblue")
abline(h = 0.5, lty = 2, col = "grey50")   # dashed line at probability 0.5
abline(v = 0,   lty = 2, col = "grey50")   # dashed line at score 0
```

## Two different questions: *explain* vs *predict*

| | **Explanatory (classical statistics)** | **Predictive (machine learning)** |
|---|---|---|
| Question | *Which factors are associated with CAD, and how strongly?* | *How accurately can we predict CAD in **new** patients?* |
| Main output | Odds ratios, 95% CI, p-values | AUC, calibration, sensitivity... on unseen data |
| Data use | Usually all data | Train/test split + cross-validation |

Researchers often mix these up. This lesson shows both, so you can see the
difference.

# Load the packages and the data

```{r packages}
library(tidyverse)
library(tidymodels)
library(gtsummary)

tidymodels_prefer()   # if two packages have a function with the same name, use the tidymodels one

heart       <- readRDS("data/heart_clean.rds")   # clean data made in Lesson 1
heart_split <- readRDS("data/heart_split.rds")   # the same train/test split as Lesson 1
heart_train <- training(heart_split)
heart_test  <- testing(heart_split)

dir.create("results", showWarnings = FALSE)       # folder to store results for Lesson 6
```

> If you get `cannot open file 'data/heart_clean.rds'`, you have not run
> Lesson 1 yet, or you did not open the course with the `.Rproj` file.

# Part A: classical (explanatory) logistic regression

`glm()` stands for *generalized linear model*. `family = binomial` tells it
to do logistic regression.

`glm()` models the probability of the **second** factor level. Our outcome
has `"Yes"` first, so to avoid confusion we make a 0/1 version where 1 = CAD.

```{r glm-fit}
heart_glm_data <- heart |>
  mutate(disease_01 = if_else(disease == "Yes", 1, 0)) |>  # 1 = CAD, 0 = no CAD
  select(-disease)                                          # drop the factor version

glm_fit <- glm(
  disease_01 ~ .,           # formula: outcome ~ predictors. The dot means "all other columns"
  data = heart_glm_data,
  family = binomial         # logistic regression
)

# glm() silently drops the 6 patients with a missing value (a "complete-case analysis").
nobs(glm_fit)   # number of patients actually used
```

These are the fitted $\beta$ values: the numbers that fill in the equation
from section 1.3. A positive $\beta$ raises the score (and the probability of
CAD); a negative $\beta$ lowers it.

```{r glm-coefficients}
round(coef(glm_fit), 3)   # coef() extracts the coefficients; round() keeps 3 decimals
```

## Odds-ratio table (publication-ready)

```{r glm-table}
tbl_regression(
  glm_fit,
  exponentiate = TRUE   # show odds ratios (exp of the coefficients) instead of log-odds
) |>
  bold_p() |>           # bold the p-values below 0.05
  bold_labels()
```

**How to read one row:** for `sex = Male`, an OR above 1 with a 95% CI that
does not include 1 means that men had higher odds of CAD than women **after
adjusting for every other variable in the model**.

> **Caution:** these are *associations*, not causes. Also, with 297 patients,
> 139 events and 17 coefficients, the model is close to the limit of what the
> data can support. A common rule of thumb asks for at least 10–20 events per
> coefficient, but this rule is crude. For proper sample-size methods see
> Riley RD, Ensor J, Snell KIE, et al. Calculating the sample size required
> for developing a clinical prediction model. *BMJ*. 2020;368:m441.
> <https://doi.org/10.1136/bmj.m441>. The R package `pmsampsize` implements
> their method.

# Part B: logistic regression as a *predictive* model (tidymodels)

Now the question is: **how well does it predict new patients?** We use the
**tidymodels** framework. You will reuse exactly the same 5 steps for random
forest, XGBoost and LightGBM:

1. **Recipe**: how to prepare the predictors.
2. **Model specification**: which algorithm.
3. **Workflow**: recipe + model glued together.
4. **Resampling (cross-validation)**: estimate performance using only the training data.
5. **Final fit**: train on all training data, then evaluate **once** on the test set.

## Step 1: the recipe (preprocessing)

A recipe lists preparation steps. The crucial point: each step **learns from
the training data only** (for example the median used to fill missing values)
and then *applies* the same thing to new data. Doing preprocessing on the full
dataset before splitting leaks test information into training. That is called
**data leakage**, and it makes results look too good.

```{r recipe}
heart_recipe <- recipe(disease ~ ., data = heart_train) |>   # outcome ~ all predictors
  step_impute_median(all_numeric_predictors()) |>  # fill missing numbers with the training median
  step_impute_mode(all_nominal_predictors()) |>    # fill missing categories with the most common one
  step_dummy(all_nominal_predictors()) |>          # turn categories into 0/1 columns ("dummy variables")
  step_zv(all_predictors()) |>                     # remove columns with zero variance (all the same value)
  step_normalize(all_numeric_predictors())         # rescale numbers to mean 0, SD 1 (needed for LASSO later)

heart_recipe
```

Want to see what the recipe produces? `prep()` learns the steps from the
training data and `bake()` applies them:

```{r recipe-peek}
heart_recipe |>
  prep() |>                 # learn medians, modes, means, SDs from the training data
  bake(new_data = NULL) |>  # apply to the training data itself (NULL = the data it learned from)
  head()                    # first 6 rows: note the new dummy columns like sex_Male
```

## Step 2: the model specification

```{r model-spec}
logistic_spec <- logistic_reg() |>   # the type of model
  set_engine("glm") |>               # which R function does the computing (glm)
  set_mode("classification")         # predicting a category, not a number

logistic_spec
```

## Step 3: the workflow

```{r workflow}
logistic_wf <- workflow() |>
  add_recipe(heart_recipe) |>    # the preprocessing
  add_model(logistic_spec)       # the model

logistic_wf
```

## Step 4: cross-validation (CV)

### Concept

Using the test set to compare or tune models would "use it up". Instead,
**10-fold cross-validation** splits the *training* data into 10 parts
(*folds*). The model is trained on 9 folds and evaluated on the remaining
fold. This repeats 10 times, so every fold is used once for evaluation. The
**average** of the 10 results estimates performance on new data.

```
Fold:   1    2    3    4    5    6    7    8    9    10
Run 1: [TEST][train..................................]
Run 2: [train][TEST][train...........................]
...
Run 10:[train..................................][TEST]
```

### Which performance measures?

| Metric | Meaning | Good value |
|--------|---------|------------|
| **ROC AUC** (C-statistic) | Discrimination: chance the model gives a random CAD patient a higher probability than a random non-CAD patient | 0.5 = coin flip, 1 = perfect |
| **Brier score** | Average squared error of the predicted probabilities (accuracy *and* calibration) | 0 = perfect. Lower is better |
| **Accuracy** | % correctly classified at a 0.5 threshold | Misleading if the outcome is rare |
| **Sensitivity** | % of CAD patients correctly flagged | Higher is better |
| **Specificity** | % of non-CAD patients correctly cleared | Higher is better |

```{r cv}
set.seed(2024)   # same seed in every lesson, so every model is tested on identical folds
heart_folds <- vfold_cv(heart_train, v = 10, strata = disease)

# The set of metrics we will compute. We reuse this in every lesson.
heart_metrics <- metric_set(roc_auc, brier_class, accuracy, sensitivity, specificity)

logistic_cv <- fit_resamples(
  logistic_wf,                                  # what to fit
  resamples = heart_folds,                      # where to fit it (the 10 folds)
  metrics   = heart_metrics,                    # what to measure
  control   = control_resamples(save_pred = TRUE)  # keep the predictions (to plot later)
)

collect_metrics(logistic_cv)   # mean = average across the 10 folds; std_err = its uncertainty
```

## Step 5: final fit and the one-time test-set evaluation

`last_fit()` trains the workflow on **all** training data and evaluates it on
the **test set**. This is the number you report as "performance on held-out
data".

```{r last-fit}
logistic_final <- last_fit(
  logistic_wf,
  split   = heart_split,    # the original split: it knows which patients are train and test
  metrics = heart_metrics
)

collect_metrics(logistic_final)   # performance on the 76 test patients
```

### Predictions for each test patient

```{r predictions}
logistic_preds <- collect_predictions(logistic_final)

logistic_preds |>
  select(disease, .pred_Yes, .pred_class) |>  # truth, predicted probability, predicted class
  head(10)
```

`.pred_Yes` is the predicted probability of CAD. `.pred_class` is "Yes" when
that probability is 0.5 or higher.

### Confusion matrix

```{r confusion, fig.height=3, fig.width=4}
logistic_preds |>
  conf_mat(truth = disease, estimate = .pred_class) |>  # cross-table of truth vs prediction
  autoplot(type = "heatmap")
```

### ROC curve

The ROC curve shows sensitivity vs (1 − specificity) at **every** possible
threshold. The closer it hugs the top-left corner, the better.

```{r roc, fig.height=4, fig.width=4}
logistic_preds |>
  roc_curve(truth = disease, .pred_Yes) |>   # compute sensitivity/specificity at all thresholds
  autoplot()
```

### The threshold is a clinical choice

0.5 is not sacred. For a screening test you may want **higher sensitivity**,
accepting more false positives. Look what happens at a threshold of 0.3:

```{r threshold}
logistic_preds |>
  mutate(
    # new predicted class: "Yes" if probability >= 0.3
    pred_class_30 = factor(if_else(.pred_Yes >= 0.3, "Yes", "No"), levels = c("Yes", "No"))
  ) |>
  summarise(
    sens_at_0.5 = sensitivity_vec(disease, .pred_class),
    spec_at_0.5 = specificity_vec(disease, .pred_class),
    sens_at_0.3 = sensitivity_vec(disease, pred_class_30),
    spec_at_0.3 = specificity_vec(disease, pred_class_30)
  )
```

# Part C: penalized logistic regression (LASSO) and your first *tuning*

**LASSO** is logistic regression with a **penalty** that shrinks weak
coefficients toward zero, and some of them to *exactly* zero. Benefits:

* It reduces **overfitting** when there are many predictors relative to events.
* It does **variable selection** automatically.

The strength of the penalty (`penalty`, also called lambda) is a
**hyperparameter**: a setting **you** choose, not something learned from the
data directly. We find a good value by **tuning**: try many values and keep
the one with the best cross-validated AUC.

```{r lasso-spec}
lasso_spec <- logistic_reg(
  penalty = tune(),   # tune() = "find the best value by cross-validation"
  mixture = 1         # 1 = LASSO, 0 = ridge, between 0 and 1 = elastic net
) |>
  set_engine("glmnet")

lasso_wf <- workflow() |>
  add_recipe(heart_recipe) |>   # same recipe (normalizing matters here: the penalty is scale-sensitive)
  add_model(lasso_spec)

# 30 penalty values from 10^-4 (almost no penalty) to 10^0 = 1 (very strong penalty)
lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 30)

lasso_tune <- tune_grid(
  lasso_wf,
  resamples = heart_folds,   # same folds as before
  grid      = lasso_grid,    # the candidate values to try
  metrics   = heart_metrics
)
```

```{r lasso-plot, fig.height=4, fig.width=8}
autoplot(lasso_tune)   # performance (y) for each penalty value (x)
```

Too little penalty overfits. Too much penalty removes useful predictors and
performance drops.

```{r lasso-best}
show_best(lasso_tune, metric = "roc_auc", n = 5)    # the 5 best penalty values

best_penalty <- select_best(lasso_tune, metric = "roc_auc")  # the single best one
best_penalty
```

> Many penalty values give almost the same AUC. If you prefer a **simpler**
> model with fewer variables, use
> `select_by_one_std_err(lasso_tune, desc(penalty), metric = "roc_auc")`.
> It picks the strongest penalty whose AUC is within one standard error of the
> best (the "1-SE rule").

## Final LASSO model

```{r lasso-final}
lasso_final_wf <- finalize_workflow(lasso_wf, best_penalty)  # put the best value into the workflow

lasso_final <- last_fit(lasso_final_wf, split = heart_split, metrics = heart_metrics)

collect_metrics(lasso_final)
```

## Which variables did LASSO keep?

Note: glmnet models the probability of the **second** outcome level ("No"),
so its coefficients point the wrong way. We flip the sign so that **positive
means higher risk of CAD**.

```{r lasso-coefs, fig.height=5, fig.width=7}
lasso_coefs <- lasso_final |>
  extract_fit_parsnip() |>                # get the fitted model out of the results
  tidy() |>                               # turn coefficients into a table
  filter(term != "(Intercept)") |>        # drop the intercept
  mutate(effect_on_CAD = -estimate)       # flip the sign: positive = higher CAD risk

lasso_coefs |> arrange(desc(abs(effect_on_CAD)))   # biggest effects first

ggplot(lasso_coefs, aes(x = effect_on_CAD, y = reorder(term, effect_on_CAD))) +
  geom_col() +
  labs(x = "Standardized coefficient (positive = higher CAD risk)", y = NULL,
       title = "LASSO coefficients (zero = variable removed)") +
  theme_minimal()
```

# Save results for the model comparison (Lesson 6)

```{r save}
# Test-set predictions, labelled with the model name
bind_rows(
  collect_predictions(logistic_final) |> mutate(model = "Logistic regression"),
  collect_predictions(lasso_final)    |> mutate(model = "LASSO logistic")
) |>
  select(model, .row, disease, .pred_Yes, .pred_class) |>  # .row = patient ID
  saveRDS("results/pred_logistic.rds")

# Cross-validation results (for the best LASSO penalty only)
bind_rows(
  collect_metrics(logistic_cv) |> mutate(model = "Logistic regression"),
  collect_metrics(lasso_tune) |>
    filter(.config == best_penalty$.config) |>   # keep only the chosen penalty
    mutate(model = "LASSO logistic")
) |>
  select(model, .metric, mean, std_err) |>
  saveRDS("results/cv_logistic.rds")
```

# Key takeaways

* **Explanatory** models give odds ratios. **Predictive** models are judged on
  **unseen data**.
* The 5-step tidymodels pattern (**recipe → model → workflow → CV → last fit**)
  is the same for every algorithm.
* Do all preprocessing **inside the recipe** to avoid data leakage.
* **Hyperparameters** (like the LASSO penalty) are tuned with
  **cross-validation**, never on the test set.

# Check your understanding

1. What is the difference between the OR table in Part A and the AUC in Part B?
2. Why is the test set used only once?
3. What does the LASSO penalty do to a weak predictor?

# Try it yourself

1. Change `mixture = 1` to `mixture = 0.5` (elastic net) and rerun Part C.
   Are more variables kept?
2. Try a threshold of 0.7 in the threshold chunk. What happens to
   sensitivity and specificity?

# References and further reading

**Methods behind this lesson**

* Riley RD, Ensor J, Snell KIE, et al. Calculating the sample size required
  for developing a clinical prediction model. *BMJ*. 2020;368:m441.
  <https://doi.org/10.1136/bmj.m441>
* Steyerberg EW, Vergouwe Y. Towards better clinical prediction models: seven
  steps for development and an ABCD for validation. *Eur Heart J*.
  2014;35(29):1925–31. <https://doi.org/10.1093/eurheartj/ehu207>.
  Covers coding of predictors, missing values and internal validation.
* Kapoor S, Narayanan A. Leakage and the reproducibility crisis in
  machine-learning-based science. *Patterns*. 2023;4(9):100804.
  <https://doi.org/10.1016/j.patter.2023.100804>. Explains why preprocessing
  must be learned from training data only, which is what the recipe does.
* Hastie T, Tibshirani R, Friedman J. *The Elements of Statistical Learning*,
  2nd ed. Springer; 2009. Section 7.10.2, "The wrong and right way to do
  cross-validation". Free PDF: <https://hastie.su.domains/ElemStatLearn/>
* Christodoulou E, Ma J, Collins GS, et al. A systematic review shows no
  performance benefit of machine learning over logistic regression for
  clinical prediction models. *J Clin Epidemiol*. 2019;110:12–22.
  <https://doi.org/10.1016/j.jclinepi.2019.02.004>
* Collins GS, Moons KGM, Dhiman P, et al. TRIPOD+AI statement. *BMJ*.
  2024;385:e078378. <https://doi.org/10.1136/bmj-2023-078378>

**Published clinical studies that used the same tidymodels workflow**

According to their abstracts, these studies built their models with
tidymodels: data splitting, resampling, preprocessing and tuning, as in this
course. Read their methods sections to see how they reported each step.

* Lu C, He Y, Chen CR, et al. Development and validation of machine learning
  models for distant metastasis of primary hepatic carcinoma: a
  population-based study. *Discov Oncol*. 2025;16:1120.
  <https://doi.org/10.1007/s12672-025-02894-5>. Eight models in tidymodels,
  with ROC/AUC, calibration, decision curve analysis and SHAP.
* Lehan E, Briand P, O'Brien E, et al. Synergistic patient factors are
  driving recent increased pediatric urgent care demand. *PLOS Digit
  Health*. 2024;3(8):e0000572. <https://doi.org/10.1371/journal.pdig.0000572>.
  tidymodels with 5-fold cross-validation and standardized predictors.
* Estrella T, Capdevila L, Alfonso C, Losilla JM. Machine learning for the
  analysis of healthy lifestyle data: scoping review and guidelines. *JMIR
  Hum Factors*. 2026;13:e78648. <https://doi.org/10.2196/78648>. Gives
  reporting guidelines and recommends tidymodels for replicable workflows.

# Next

Open **`03_random_forest.Rmd`**.

