---
title: "Lesson 3 — Decision trees and random forests"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The concept

## Step 1: a single decision tree

A **decision tree** asks a series of yes/no questions, like a clinical
algorithm:

```
Is chest pain asymptomatic?
├── yes → Is thal a reversible defect? → yes → high risk
└── no  → Are there 0 diseased vessels on fluoroscopy?  → yes → low risk
```

The computer chooses each question (variable + cut-off) to best separate CAD
from no CAD.

* **Good:** easy to understand, captures interactions and non-linear effects,
  no need to rescale variables.
* **Bad:** **unstable**. Change a few patients and you get a very different
  tree. A deep tree **overfits** (memorises the training data).

## Step 2: a random forest = many decorrelated trees that vote

A **random forest** (Breiman, 2001) grows hundreds of trees, each slightly
different, and **averages their votes**:

1. **Bootstrap (bagging):** each tree is trained on a random resample of the
   patients (drawn *with replacement*).
2. **Random feature subset:** at each split, a tree may only choose from a
   random subset of `mtry` variables. This stops all trees from looking the
   same.
3. **Average:** the predicted probability is the share of trees voting "CAD".

Averaging many noisy-but-different trees cancels out their errors. That is the
"wisdom of crowds".

## The hyperparameters you will tune

| Hyperparameter | Meaning | Effect |
|----------------|---------|--------|
| `mtry` | Number of variables each split can choose from | Small = more diverse trees. Large = stronger individual trees |
| `min_n` | Minimum patients in a node to split it further | Small = deeper, more complex trees. Large = simpler trees |
| `trees` | Number of trees | More is better up to a point (500–1000 is usually enough). Not usually tuned |

# Load packages and data

```{r packages}
library(tidyverse)
library(tidymodels)
tidymodels_prefer()

heart_split <- readRDS("data/heart_split.rds")   # same split as all other lessons
heart_train <- training(heart_split)
heart_test  <- testing(heart_split)

set.seed(2024)                                    # same folds as Lesson 2
heart_folds <- vfold_cv(heart_train, v = 10, strata = disease)

heart_metrics <- metric_set(roc_auc, brier_class, accuracy, sensitivity, specificity)
```

# The recipe: trees need less preprocessing

Trees split on cut-offs ("is age > 55?"), so they **don't need** dummy
variables or normalization. The `ranger` package handles categories directly.
We only fill in missing values.

```{r recipe}
rf_recipe <- recipe(disease ~ ., data = heart_train) |>
  step_impute_median(all_numeric_predictors()) |>   # fill missing numbers (training median)
  step_impute_mode(all_nominal_predictors())        # fill missing categories (most frequent)
```

# Warm-up: one decision tree, to see the idea

```{r single-tree, fig.height=6, fig.width=9}
tree_spec <- decision_tree(tree_depth = 3) |>  # limit to 3 levels of questions, to keep it readable
  set_engine("rpart") |>                       # rpart = recursive partitioning (comes with R)
  set_mode("classification")

tree_fit <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(tree_spec) |>
  fit(data = heart_train)                      # fit() trains on the data given

# Draw the tree. extract_fit_engine() gives the underlying rpart object.
tree_engine <- extract_fit_engine(tree_fit)
plot(tree_engine, uniform = TRUE, margin = 0.1)   # draw the branches
text(tree_engine, use.n = TRUE, pretty = 0, cex = 0.7)   # labels; use.n = counts (Yes/No); pretty = 0 = full category names
```

Read it from the top: at each split, patients for whom the condition is
**true go left**. The leaves show the predicted class and the number of
`Yes/No` training patients that ended up there.

# Random forest: the 5 tidymodels steps

## Model specification with hyperparameters to tune

```{r rf-spec}
rf_spec <- rand_forest(
  mtry  = tune(),     # to be tuned
  min_n = tune(),     # to be tuned
  trees = 500         # fixed: 500 trees
) |>
  set_engine(
    "ranger",
    importance = "permutation",  # compute variable importance (explained below)
    seed = 2024                  # makes ranger's internal randomness reproducible
  ) |>
  set_mode("classification")

rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)
```

## Tuning with a grid search + cross-validation

We try 5 values of `mtry` × 5 values of `min_n` = **25 combinations**. Each
is evaluated with 10-fold CV, so that is 250 forests. This may take a minute.

```{r rf-tune}
rf_grid <- grid_regular(
  mtry(range = c(1, 10)),    # we have 13 predictors. Try 1 to 10 per split
  min_n(range = c(2, 30)),   # minimum node size from 2 to 30 patients
  levels = 5                 # 5 values of each, evenly spaced
)

rf_grid   # look at the combinations we will try

rf_tune <- tune_grid(
  rf_wf,
  resamples = heart_folds,
  grid      = rf_grid,
  metrics   = heart_metrics
)
```

```{r rf-tune-plot, fig.height=5, fig.width=9}
autoplot(rf_tune) +                                       # one line per min_n value. x-axis = mtry
  scale_x_continuous(breaks = sort(unique(rf_grid$mtry))) +   # ticks only at the mtry values we tried
  labs(x = "Variables tried at each split (mtry)",
       colour = "Minimum node size (min_n)", shape = "Minimum node size (min_n)")
```

The ticks on the x-axis are exactly the `mtry` values in the grid (1, 3, 5,
7 and 10 variables), so you can read each point straight off the axis.


```{r rf-best}
show_best(rf_tune, metric = "roc_auc", n = 5)
best_rf <- select_best(rf_tune, metric = "roc_auc")
best_rf
```

> **Tip:** if the best value sits at the **edge** of the grid (for example
> `min_n = 30`), widen the range and tune again. The optimum might lie beyond.
> (`mtry` cannot go below 1. A best `mtry` of 1 just means very random,
> very diverse trees worked well on this small dataset.)

## Final model and test-set evaluation

```{r rf-final}
rf_final_wf <- finalize_workflow(rf_wf, best_rf)   # plug in the best mtry and min_n

rf_final <- last_fit(rf_final_wf, split = heart_split, metrics = heart_metrics)

collect_metrics(rf_final)   # performance on the untouched test set
```

```{r rf-roc, fig.height=4, fig.width=4}
collect_predictions(rf_final) |>
  roc_curve(truth = disease, .pred_Yes) |>
  autoplot()
```

# Which variables matter? (variable importance)

A forest has no odds ratios. Instead we use **permutation importance**:
shuffle one variable's values at random, which destroys its information, and
measure how much the prediction accuracy drops. A big drop means an important
variable.

```{r rf-importance, fig.height=4, fig.width=7}
rf_importance <- rf_final |>
  extract_fit_engine() |>            # the ranger model inside the workflow
  pluck("variable.importance") |>    # pluck() pulls one element out of a list
  enframe(name = "variable", value = "importance")   # named numbers to a two-column table

ggplot(rf_importance, aes(x = importance, y = reorder(variable, importance))) +
  geom_col(fill = "forestgreen") +
  labs(x = "Permutation importance (drop in accuracy when shuffled)", y = NULL,
       title = "Random forest variable importance") +
  theme_minimal()
```

> **Interpretation caution:** importance says **how much** the model relies on
> a variable, not the **direction** (risk up or down), and not causation.
> Correlated variables can share or "steal" importance from each other.

# Save results for Lesson 6

```{r save}
collect_predictions(rf_final) |>
  mutate(model = "Random forest") |>
  select(model, .row, disease, .pred_Yes, .pred_class) |>  # .row = patient ID
  saveRDS("results/pred_rf.rds")

collect_metrics(rf_tune) |>
  filter(.config == best_rf$.config) |>
  mutate(model = "Random forest") |>
  select(model, .metric, mean, std_err) |>
  saveRDS("results/cv_rf.rds")
```

# Key takeaways

* A **decision tree** is interpretable but unstable. A **random forest**
  averages many randomized trees and is much more stable and accurate.
* Key hyperparameters: `mtry` and `min_n`. Tune them with **CV**.
* Trees need little preprocessing: no dummies, no scaling.
* Use **permutation importance** to see what the model relies on.

# Check your understanding

1. What two sources of randomness make the trees in a forest different?
2. Why does a single deep tree overfit?
3. What does permutation importance **not** tell you?

# Try it yourself

1. Change `trees = 500` to `trees = 100` and then `2000`. Does test AUC change
   much?

   > **Answer:** no. In our run the test AUC was about 0.86 with 100, 500 and
   > 2000 trees (it moved only in the second decimal). After a few hundred
   > trees the average vote stabilises. Adding more trees **does not cause
   > overfitting** in a random forest; it only costs computing time. That is
   > why `trees` is usually fixed (500–1000) rather than tuned.

2. Make the single tree deeper. Change `tree_depth = 3` to `5`. Does the tree
   change?

   > **Answer:** surprisingly, no: you get the same 3-level tree.
   > `tree_depth` is a **maximum**, not a target. `rpart` stops earlier when
   > its other two stopping rules say a further split isn't worth it:
   >
   > * `cost_complexity` (default 0.01): a split is kept only if it improves
   >   the overall fit by at least 1%. Nodes that are already fairly "pure"
   >   (mostly Yes or mostly No) are not split further.
   > * `min_n` (default 20): a node needs at least 20 patients before `rpart`
   >   will try to split it. Small nodes stay as leaves.
   >
   > To see a deeper tree, relax all three settings:

   ```{r deeper-tree, fig.height=7, fig.width=11}
   tree_spec_deep <- decision_tree(
     tree_depth      = 5,       # allow up to 5 levels of questions
     cost_complexity = 0.001,   # accept much smaller improvements per split
     min_n           = 5        # allow splitting nodes with as few as 5 patients
   ) |>
     set_engine("rpart") |>
     set_mode("classification")

   tree_fit_deep <- workflow() |>
     add_recipe(rf_recipe) |>
     add_model(tree_spec_deep) |>
     fit(data = heart_train)

   tree_engine_deep <- extract_fit_engine(tree_fit_deep)
   plot(tree_engine_deep, uniform = TRUE, margin = 0.05)
   text(tree_engine_deep, use.n = TRUE, pretty = 0, cex = 0.6)
   ```

   This tree has 5 levels and many more leaves. It is much harder to read, and
   its small leaves (a handful of patients each) are a clear sign of
   **overfitting**. That is exactly why random forests average many trees
   instead of trusting one deep tree.

# Next

Open **`04_xgboost.Rmd`**.
