---
title: "Lesson 4 — Gradient boosting with XGBoost"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The concept

## Boosting vs random forest

* **Random forest:** many **deep, independent** trees grown **in parallel**,
  then averaged.
* **Boosting:** many **small** trees grown **one after another**. Each new
  tree focuses on the errors the previous trees still make.

Analogy: a team of students correcting an essay. Student 1 writes a rough
draft. Student 2 fixes only what's still wrong. Student 3 fixes what's still
wrong after that, and so on. Each correction is small, but together they
produce a very good essay.

**"Gradient"** boosting means each tree is fitted to the *gradient* of the
loss: in plain words, to the direction in which the current predictions are
most wrong.

**XGBoost** ("eXtreme Gradient Boosting"; Chen & Guestrin, 2016) is a fast,
regularized implementation that has won many prediction competitions on
tabular data.

## The hyperparameters that matter

| Hyperparameter (tidymodels name) | XGBoost name | Meaning | Typical range |
|---|---|---|---|
| `trees` | `nrounds` | Number of trees added in sequence | 100–2000 |
| `learn_rate` | `eta` | How much each tree's correction counts. Small = slow but careful | 0.001–0.3 |
| `tree_depth` | `max_depth` | Size of each tree. Small = simple interactions | 1–6 |
| `min_n` | `min_child_weight` | Minimum data in a leaf. Larger = more conservative | 1–30 |
| `sample_size` | `subsample` | Share of patients used for each tree (adds randomness) | 0.5–1 |

**Trade-off:** a small `learn_rate` needs many `trees`. Boosting **can
overfit** if it runs too long with too high a learning rate, which is why we
tune these together with cross-validation.

# Load packages and data

```{r packages}
library(tidyverse)
library(tidymodels)
tidymodels_prefer()

heart_split <- readRDS("data/heart_split.rds")
heart_train <- training(heart_split)
heart_test  <- testing(heart_split)

set.seed(2024)
heart_folds <- vfold_cv(heart_train, v = 10, strata = disease)

heart_metrics <- metric_set(roc_auc, brier_class, accuracy, sensitivity, specificity)
```

# The recipe: XGBoost needs numbers only

XGBoost only accepts numbers, so categories must become **dummy (0/1)
columns**. It **handles missing numeric values by itself**: at every split it
learns which direction missing values should go. So we only fill in missing
*categories* (before making dummies) and leave missing numbers as they are.

```{r recipe}
xgb_recipe <- recipe(disease ~ ., data = heart_train) |>
  step_impute_mode(all_nominal_predictors()) |>   # fill missing categories (e.g. thal)
  step_dummy(all_nominal_predictors())            # categories to 0/1 columns. No scaling needed for trees
```

# Model specification with tuning

```{r xgb-spec}
xgb_spec <- boost_tree(
  trees       = tune(),
  tree_depth  = tune(),
  learn_rate  = tune(),
  min_n       = tune(),
  sample_size = 0.8        # fixed: each tree sees a random 80% of patients
) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_wf <- workflow() |>
  add_recipe(xgb_recipe) |>
  add_model(xgb_spec)
```

## Set sensible search ranges

The default ranges are very wide, so we narrow them to values that make sense
for small tabular data.

```{r xgb-params}
xgb_params <- xgb_wf |>
  extract_parameter_set_dials() |>             # the list of parameters marked tune()
  update(                                      # change their search ranges:
    trees      = trees(range = c(100, 1000)),
    tree_depth = tree_depth(range = c(1, 6)),
    learn_rate = learn_rate(range = c(-3, -0.5)),  # on log10 scale: 0.001 to 0.32
    min_n      = min_n(range = c(2, 30))
  )

xgb_params
```

## Space-filling grid

With 4 hyperparameters, a regular grid of 5 values each would be
5^4 = 625 combinations. That is too many. A **space-filling grid** spreads a
smaller number of combinations (here 30) evenly across the whole search space.

```{r xgb-tune}
set.seed(2024)
xgb_grid <- grid_space_filling(xgb_params, size = 30)

head(xgb_grid)   # the first few combinations

xgb_tune <- tune_grid(
  xgb_wf,
  resamples = heart_folds,
  grid      = xgb_grid,
  metrics   = heart_metrics
)
```

```{r xgb-tune-plot, fig.height=6, fig.width=10}
autoplot(xgb_tune, metric = "roc_auc")   # AUC against each hyperparameter
```

```{r xgb-best}
show_best(xgb_tune, metric = "roc_auc", n = 5)
best_xgb <- select_best(xgb_tune, metric = "roc_auc")
best_xgb
```

# Final model and test-set evaluation

```{r xgb-final}
xgb_final_wf <- finalize_workflow(xgb_wf, best_xgb)

xgb_final <- last_fit(xgb_final_wf, split = heart_split, metrics = heart_metrics)

collect_metrics(xgb_final)
```

```{r xgb-roc, fig.height=4, fig.width=4}
collect_predictions(xgb_final) |>
  roc_curve(truth = disease, .pred_Yes) |>
  autoplot()
```

# Explaining the model

## Built-in importance (gain)

**Gain** = how much a variable improved the predictions, summed over all the
splits that used it.

```{r xgb-importance, fig.height=5, fig.width=7}
xgb_booster <- extract_fit_engine(xgb_final)    # the raw XGBoost model

xgb_importance <- xgboost::xgb.importance(model = xgb_booster)  # package::function() calls a function without library()

ggplot(xgb_importance, aes(x = Gain, y = reorder(Feature, Gain))) +
  geom_col(fill = "steelblue") +
  labs(x = "Importance (gain)", y = NULL, title = "XGBoost variable importance") +
  theme_minimal()
```

## SHAP values: importance **with direction**, for every patient

**SHAP values** (SHapley Additive exPlanations; Lundberg & Lee, 2017) split
each patient's prediction into contributions from each variable:

`prediction (log-odds) = baseline + SHAP(age) + SHAP(sex) + ... + SHAP(thal)`

* **Positive SHAP:** that variable pushed **this patient's** risk **up**.
* **Negative SHAP:** it pushed the risk **down**.

They are widely used in ML papers to open up the "black box". XGBoost
computes them exactly for tree models.

```{r shap-compute}
# 1. Apply the trained recipe to the test patients, to get the exact columns XGBoost saw
xgb_fitted_wf <- extract_workflow(xgb_final)           # the trained workflow
test_x <- xgb_fitted_wf |>
  extract_recipe() |>                                  # its trained recipe
  bake(new_data = heart_test, all_predictors())        # apply it; keep predictors only

# 2. Ask XGBoost for SHAP contributions (predcontrib = TRUE)
shap_matrix <- predict(xgb_booster, newdata = as.matrix(test_x), predcontrib = TRUE)

# 3. Put them in a long table: one row per patient × variable
shap_long <- as_tibble(shap_matrix) |>
  select(-`(Intercept)`) |>                      # drop the baseline column
  mutate(patient = row_number()) |>
  pivot_longer(-patient, names_to = "variable", values_to = "shap") |>
  left_join(                                     # add each patient's actual value of the variable
    test_x |>
      mutate(patient = row_number()) |>
      pivot_longer(-patient, names_to = "variable", values_to = "value"),
    by = c("patient", "variable")
  ) |>
  group_by(variable) |>
  mutate(value_scaled = percent_rank(value)) |>  # 0 = lowest value, 1 = highest (for colouring)
  ungroup()

head(shap_long)
```

### SHAP summary ("beeswarm") plot

Each dot is one patient. **Position** shows how much the variable pushed that
patient's risk up (right) or down (left). **Colour** shows whether the
patient's value was high (red) or low (blue).

```{r shap-beeswarm, fig.height=6, fig.width=8}
shap_order <- shap_long |>
  group_by(variable) |>
  summarise(mean_abs = mean(abs(shap))) |>       # average size of the effect
  arrange(mean_abs) |>
  pull(variable)                                 # variable names, least to most important

ggplot(shap_long, aes(x = shap, y = factor(variable, levels = shap_order), colour = value_scaled)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_jitter(height = 0.2, width = 0, alpha = 0.8) +   # spread dots vertically so they don't overlap
  scale_colour_gradient(low = "blue", high = "red", name = "Variable value\n(low → high)") +
  labs(x = "SHAP value (contribution to log-odds of CAD)", y = NULL,
       title = "SHAP summary plot, test patients") +
  theme_minimal()
```

**Example reading:** for `n_vessels`, red dots (more diseased vessels) sit on
the right, so more vessels push the predicted risk of CAD up. That matches
clinical knowledge, a good sanity check.

### SHAP dependence plot for one variable

```{r shap-dependence, fig.height=4, fig.width=6}
shap_long |>
  filter(variable == "max_hr") |>
  ggplot(aes(x = value, y = shap)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess") +   # trend line
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Maximum heart rate (bpm)", y = "SHAP value",
       title = "How max heart rate changes the predicted risk") +
  theme_minimal()
```

This shows the **shape** of the relationship, which may be non-linear,
something plain logistic regression assumes away.

# Save results for Lesson 6

```{r save}
collect_predictions(xgb_final) |>
  mutate(model = "XGBoost") |>
  select(model, .row, disease, .pred_Yes, .pred_class) |>  # .row = patient ID
  saveRDS("results/pred_xgb.rds")

collect_metrics(xgb_tune) |>
  filter(.config == best_xgb$.config) |>
  mutate(model = "XGBoost") |>
  select(model, .metric, mean, std_err) |>
  saveRDS("results/cv_xgb.rds")
```

# Key takeaways

* Boosting builds **small trees in sequence**, each correcting the previous
  errors.
* The most important hyperparameters are `learn_rate`, `trees` and
  `tree_depth`. Tune them **together**.
* XGBoost needs numeric input (dummies) but handles missing numbers natively.
* **SHAP values** give per-patient, direction-aware explanations.
* On small datasets like this one, boosting often does **not** beat
  logistic regression. That is a real finding, not a failure.

# Check your understanding

1. What is the key difference between how a random forest and boosting build
   their trees?
2. Why does a smaller learning rate need more trees?
3. What does a positive SHAP value mean for a patient?

# Try it yourself

1. Change `size = 30` to `size = 10` in the grid. Does the best AUC change?
2. Make a SHAP dependence plot for `age` instead of `max_hr`.

# Next

Open **`05_lightgbm.Rmd`**.
