---
title: "✏️ Machine-learning prediction of malignancy in breast fine-needle aspirates"
subtitle: "Research project template (worked example: Breast Cancer Wisconsin Diagnostic data)"
author: "✏️ Your Name, Affiliation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
# code_folding: hide (in the header) hides code in the HTML report behind
# "Code" buttons, so the report reads like a paper. Change it to "show" to see all code.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<!--
=====================================================================
HOW TO USE THIS TEMPLATE  (this block is an HTML comment: it never
appears in the knitted report)

1. Save a copy under a new name (File > Save As), e.g. "my_study.Rmd".
2. Search for the pencil ✏️ symbol. Every place marked ✏️ is something
   you edit. Everything else can usually stay as it is.
3. The ONLY strict requirement: the "Data preparation" section must create
   a data frame called `study_data` that contains
      * a column called `outcome`: a factor with levels c("Yes", "No"),
        where "Yes" is the event you want to predict, and
      * the predictor columns ONLY (no IDs, no dates, nothing measured
        after the outcome, nothing that "contains the answer").
4. Run chunk by chunk first (green ▶). When everything works, Knit.
=====================================================================
-->

# Background ✏️

*State what is known, what is unknown, and why prediction matters here.*

Breast cancer diagnosis often relies on fine-needle aspiration (FNA)
cytology. Quantitative features of cell nuclei, measured from digitized FNA
images, may help separate malignant from benign masses. This study compares
classical and machine-learning (ML) models for this task.

# Objectives ✏️

* **Primary:** develop and internally validate models that predict malignancy
  from nuclear features of FNA images.
* **Secondary:** compare logistic regression, LASSO, random forest, XGBoost
  and LightGBM on discrimination, calibration and clinical utility, and
  identify the most influential features.

*Write these **before** analysing the data, and ideally register them.*

# Methods

## Data source ✏️

Breast Cancer Wisconsin (Diagnostic) dataset, 569 FNA samples from the
University of Wisconsin Hospitals (Wolberg, Street, Mangasarian). Public
via the UCI Machine Learning Repository:

* Wolberg W, Mangasarian O, Street N, Street W. *Breast Cancer Wisconsin
  (Diagnostic)* [Dataset]. UCI Machine Learning Repository, 1993.
  <https://doi.org/10.24432/C5DW2B>

## Outcome ✏️

Histologically confirmed **malignant** (`Yes`) vs **benign** (`No`) mass.

## Predictors ✏️

Ten nuclear characteristics (radius, texture, perimeter, area, smoothness,
compactness, concavity, concave points, symmetry, fractal dimension), each
summarised as the **mean**, **standard error** and **worst** (mean of the 3
largest values) per image: 30 predictors in total.

## Data preparation

```{r packages}
library(tidyverse)
library(tidymodels)
library(bonsai)
library(gtsummary)
tidymodels_prefer()

dir.create("data", showWarnings = FALSE)
```

```{r data-import}
# ✏️ ---- REPLACE THIS CHUNK WITH THE IMPORT OF YOUR OWN DATA -----------------
# Examples:
#   my_raw <- read_csv("data/my_file.csv")            # CSV file
#   my_raw <- readxl::read_excel("data/my_file.xlsx") # Excel file
#   my_raw <- haven::read_sav("data/my_file.sav")     # SPSS file

data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data"

measures <- c("radius", "texture", "perimeter", "area", "smoothness",
              "compactness", "concavity", "concave_points", "symmetry",
              "fractal_dimension")

# The file has: ID, diagnosis, then the 10 means, 10 standard errors, 10 "worst" values
column_names <- c("id", "diagnosis",
                  paste0(measures, "_mean"),   # paste0() glues text: "radius" + "_mean" = "radius_mean"
                  paste0(measures, "_se"),
                  paste0(measures, "_worst"))

raw_data <- read_csv(data_url, col_names = column_names)
write_csv(raw_data, "data/wdbc_raw.csv")   # keep a local untouched copy
```

```{r data-cleaning}
# ✏️ ---- ADAPT: build `study_data` with `outcome` (Yes/No) + predictors only ----
study_data <- raw_data |>
  mutate(outcome = factor(if_else(diagnosis == "M", "Yes", "No"),   # M = malignant = the event
                          levels = c("Yes", "No"))) |>              # "Yes" MUST be the first level
  select(-id, -diagnosis)                                           # drop ID and the original outcome

# Safety checks: these stop the knitting with a clear message if something is wrong
stopifnot("outcome" %in% names(study_data))
stopifnot(identical(levels(study_data$outcome), c("Yes", "No")))

glimpse(study_data)
```

## Statistical analysis

*(This paragraph is written to match the code below. Update it if you change
the code.)*

Data were randomly split into a development set (75%) and a held-out test
set (25%), stratified by outcome. Missing values were imputed (median for
continuous, mode for categorical predictors) **within** each resampling step
to prevent information leakage. Five models were developed: unpenalized
logistic regression, LASSO logistic regression, random forest (`ranger`),
XGBoost and LightGBM. Hyperparameters were tuned by 10-fold cross-validation
on the development set using a space-filling grid of 20 candidates per model,
selecting the value with the highest ROC AUC. Final models were refitted on
the full development set and evaluated once on the test set.
**Discrimination** was assessed with ROC AUC, **overall accuracy of
probabilities** with the Brier score, and **calibration** with calibration
plots. 95% CIs were obtained by 1,000 bootstrap resamples of the test set.
**Clinical utility** was assessed with decision curve analysis. Feature
importance for the best model was estimated by permutation on the test set.
Analyses used R `r getRversion()` and tidymodels. Reporting follows TRIPOD+AI.

```{r split}
set.seed(2024)                                                     # ✏️ any number, but report it
data_split <- initial_split(study_data, prop = 0.75, strata = outcome)
train_data <- training(data_split)
test_data  <- testing(data_split)

set.seed(2024)
cv_folds <- vfold_cv(train_data, v = 10, strata = outcome)        # ✏️ v = 5 is faster

my_metrics <- metric_set(roc_auc, brier_class, accuracy, sensitivity, specificity)
```

# Results

## Study population

There were `r nrow(study_data)` samples, of which
`r sum(study_data$outcome == "Yes")`
(`r round(100 * mean(study_data$outcome == "Yes"), 1)`%) had the outcome.
The development set had `r nrow(train_data)` and the test set
`r nrow(test_data)` samples.

```{r table1}
# ✏️ choose which variables to show in Table 1 (here: only the 10 "_mean" features)
study_data |>
  select(outcome, ends_with("_mean")) |>
  tbl_summary(by = outcome) |>
  add_overall() |>
  add_p() |>
  modify_caption("**Table 1. Characteristics by outcome**")
```

```{r missing-check}
# Number of missing values per variable (only those with any)
missing_counts <- colSums(is.na(study_data))
missing_counts[missing_counts > 0]
```

## Model development

```{r model-specs}
# ---- Preprocessing recipes -------------------------------------------------
# For linear models and XGBoost: impute, dummy-code, remove constant columns, standardize
recipe_norm <- recipe(outcome ~ ., data = train_data) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

# For random forest and LightGBM: only impute (they handle categories themselves)
recipe_basic <- recipe(outcome ~ ., data = train_data) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors())

# ---- Model specifications ------------------------------------------------
logistic_spec <- logistic_reg() |> set_engine("glm")

lasso_spec <- logistic_reg(penalty = tune(), mixture = 1) |> set_engine("glmnet")

rf_spec <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) |>
  set_engine("ranger", seed = 2024) |>
  set_mode("classification")

xgb_spec <- boost_tree(trees = tune(), tree_depth = tune(), learn_rate = tune(),
                       min_n = tune(), sample_size = 0.8) |>
  set_engine("xgboost") |>
  set_mode("classification")

lgb_spec <- boost_tree(trees = tune(), learn_rate = tune(), min_n = tune()) |>
  set_engine("lightgbm", num_leaves = tune(), num_threads = 1, verbose = -1) |>
  set_mode("classification")

# ---- Pair each recipe with its model in a "workflow set" ------------------
# cross = FALSE pairs them by position (1st recipe + 1st model, 2nd + 2nd, ...)
model_set <- workflow_set(
  preproc = list(norm = recipe_norm, norm = recipe_norm, basic = recipe_basic,
                 norm = recipe_norm, basic = recipe_basic),
  models  = list(logistic = logistic_spec, lasso = lasso_spec, rf = rf_spec,
                 xgb = xgb_spec, lgb = lgb_spec),
  cross   = FALSE
)

# Sensible search ranges for the boosting models (see Lessons 4 and 5)
xgb_params <- workflow(recipe_norm, xgb_spec) |>
  extract_parameter_set_dials() |>
  update(trees = trees(c(100, 1000)), tree_depth = tree_depth(c(1, 6)),
         learn_rate = learn_rate(c(-3, -0.5)), min_n = min_n(c(2, 30)))

lgb_params <- workflow(recipe_basic, lgb_spec) |>
  extract_parameter_set_dials() |>
  update(trees = trees(c(100, 1000)), learn_rate = learn_rate(c(-3, -0.5)),
         min_n = min_n(c(5, 40)), num_leaves = num_leaves(c(4, 31)))

model_set <- model_set |>
  option_add(param_info = xgb_params, id = "norm_xgb") |>
  option_add(param_info = lgb_params, id = "basic_lgb")

model_set   # one row per model ("wflow_id")
```

```{r tuning}
# Tune every model with the same folds. This is the slow step (a few minutes).
# A model with nothing to tune (plain logistic) is simply cross-validated.
model_results <- workflow_map(
  model_set,
  fn        = "tune_grid",
  resamples = cv_folds,
  grid      = 20,            # ✏️ 20 candidate settings per model
  metrics   = my_metrics,
  seed      = 2024,
  verbose   = FALSE          # set to TRUE to see progress messages
)
```

```{r cv-ranking, fig.height=4, fig.width=7}
model_labels <- c(norm_logistic = "Logistic regression", norm_lasso = "LASSO logistic",
                  basic_rf = "Random forest", norm_xgb = "XGBoost", basic_lgb = "LightGBM")

rank_results(model_results, rank_metric = "roc_auc", select_best = TRUE) |>
  filter(.metric == "roc_auc") |>
  transmute(Model = model_labels[wflow_id], `CV AUC` = mean, SE = std_err, Rank = rank) |>
  knitr::kable(digits = 3, caption = "Table 2. Cross-validated AUC (best setting per model)")

autoplot(model_results, rank_metric = "roc_auc", metric = "roc_auc", select_best = TRUE) +
  labs(title = "Cross-validated AUC, best configuration of each model")
```

## Performance on the held-out test set

```{r final-fits}
# For each model: take its best hyperparameters, refit on all training data, test once
final_fits <- map(model_results$wflow_id, function(id) {
  best_params <- model_results |>
    extract_workflow_set_result(id) |>
    select_best(metric = "roc_auc")
  model_results |>
    extract_workflow(id) |>
    finalize_workflow(best_params) |>
    last_fit(split = data_split, metrics = my_metrics)
}) |>
  set_names(model_results$wflow_id)   # name each result by its model ID

# Bootstrap 95% CIs of the test-set metrics (int_pctl = percentile intervals)
test_results <- imap(final_fits, function(fit, id) {      # imap gives each element AND its name
  set.seed(2024)
  int_pctl(fit, times = 1000) |> mutate(model = model_labels[id])
}) |>
  list_rbind()

test_results |>
  filter(.metric %in% c("roc_auc", "brier_class", "sensitivity", "specificity")) |>
  mutate(value = sprintf("%.3f (%.3f–%.3f)", .estimate, .lower, .upper)) |>
  select(model, .metric, value) |>
  pivot_wider(names_from = .metric, values_from = value) |>
  knitr::kable(caption = "Table 3. Test-set performance, estimate (95% bootstrap CI)")
```

```{r test-preds}
test_preds <- imap(final_fits, function(fit, id) {
  collect_predictions(fit) |> mutate(model = model_labels[id])
}) |>
  list_rbind() |>
  mutate(model = factor(model, levels = unname(model_labels)))   # keep models in a fixed order
```

```{r roc-plot, fig.height=5, fig.width=6}
test_preds |>
  group_by(model) |>
  roc_curve(truth = outcome, .pred_Yes) |>
  autoplot() +
  labs(title = "Figure 1. ROC curves, test set")
```

```{r calibration-plot, fig.height=5, fig.width=9}
test_preds |>
  group_by(model) |>
  mutate(risk_group = ntile(.pred_Yes, 5)) |>     # ✏️ use 10 groups if your test set is large
  group_by(model, risk_group) |>
  summarise(predicted = mean(.pred_Yes), observed = mean(outcome == "Yes"),
            n = n(), .groups = "drop") |>
  ggplot(aes(predicted, observed)) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue", size = 2) +
  facet_wrap(~ model) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Predicted probability", y = "Observed proportion",
       title = "Figure 2. Calibration, test set") +
  theme_minimal()
```

```{r dca, fig.height=5, fig.width=8}
thresholds <- tibble(threshold = seq(0.05, 0.80, by = 0.01))
prevalence <- mean(test_data$outcome == "Yes")

model_curves <- test_preds |>
  cross_join(thresholds) |>
  group_by(model, threshold) |>
  summarise(tp_rate = sum(.pred_Yes >= threshold & outcome == "Yes") / n(),   # true positives / n
            fp_rate = sum(.pred_Yes >= threshold & outcome == "No")  / n(),   # false positives / n
            .groups = "drop") |>
  mutate(net_benefit = tp_rate - fp_rate * threshold / (1 - threshold),
         model = as.character(model))

treat_all <- thresholds |>
  mutate(model = "Treat all",
         net_benefit = prevalence - (1 - prevalence) * threshold / (1 - threshold))

bind_rows(model_curves, treat_all) |>
  ggplot(aes(threshold, net_benefit, colour = model)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0) +
  coord_cartesian(ylim = c(-0.05, prevalence + 0.05)) +
  labs(x = "Risk threshold", y = "Net benefit", colour = NULL,
       title = "Figure 3. Decision curve analysis, test set") +
  theme_minimal()
```

## Feature importance of the best model

**Permutation importance** works for *any* model: shuffle one predictor and
measure how much the test AUC drops.

```{r permutation-importance, fig.height=6, fig.width=7}
best_id <- test_results |>
  filter(.metric == "roc_auc") |>
  slice_max(.estimate, n = 1, with_ties = FALSE) |>
  pull(model)
best_id <- names(model_labels)[model_labels == best_id]   # turn the label back into the ID
best_workflow <- extract_workflow(final_fits[[best_id]])  # the trained workflow

permutation_importance <- function(fitted_wf, data, n_repeats = 10) {
  # AUC with the original data
  baseline <- roc_auc_vec(data$outcome, predict(fitted_wf, data, type = "prob")$.pred_Yes)
  predictors <- setdiff(names(data), "outcome")
  map(predictors, function(var) {
    drops <- map_dbl(1:n_repeats, function(r) {
      shuffled <- data
      shuffled[[var]] <- sample(shuffled[[var]])   # destroy this variable's information
      baseline - roc_auc_vec(shuffled$outcome,
                             predict(fitted_wf, shuffled, type = "prob")$.pred_Yes)
    })
    tibble(variable = var, importance = mean(drops))
  }) |>
    list_rbind()
}

set.seed(2024)
importance <- permutation_importance(best_workflow, test_data)

importance |>
  slice_max(importance, n = 15) |>                    # top 15 predictors
  ggplot(aes(importance, reorder(variable, importance))) +
  geom_col(fill = "steelblue") +
  labs(x = "Drop in test AUC when shuffled", y = NULL,
       title = paste("Figure 4. Permutation importance:", model_labels[best_id])) +
  theme_minimal()
```

## Summary of main result

```{r summary-numbers}
auc_row <- test_results |>
  filter(.metric == "roc_auc") |>
  slice_max(.estimate, n = 1, with_ties = FALSE)
logit_auc <- test_results |>
  filter(.metric == "roc_auc", model == "Logistic regression")
```

The best test-set discrimination was achieved by **`r auc_row$model`**
(AUC `r sprintf("%.3f", auc_row$.estimate)`, 95% CI
`r sprintf("%.3f", auc_row$.lower)`–`r sprintf("%.3f", auc_row$.upper)`),
compared with `r sprintf("%.3f", logit_auc$.estimate)` for unpenalized
logistic regression.

# Discussion ✏️

*Prompts. Replace each with your own text:*

* **Principal findings.** What did you find, in one or two sentences?
* **Comparison with the literature.** Do ML models outperform logistic
  regression here? Is that consistent with prior studies?
* **Calibration and clinical utility.** Are the probabilities trustworthy? At
  which thresholds is the model useful?
* **Strengths.** Public data, pre-specified analysis, leakage-free
  resampling, full code available.
* **Limitations.** Sample size, single centre, internal validation only, no
  external validation, possible spectrum bias, missing data assumptions.
* **Implications and next steps.** External validation? Prospective study?

# Reproducibility

```{r session-info}
sessionInfo()
```

# References ✏️

1. Collins GS, et al. TRIPOD+AI statement: updated guidance for reporting
   clinical prediction models that use regression or machine learning
   methods. *BMJ*. 2024;385:e078378.
2. Kuhn M, Silge J. *Tidy Modeling with R*. O'Reilly; 2022.
   <https://www.tmwr.org>
3. Wolberg W, et al. Breast Cancer Wisconsin (Diagnostic). UCI Machine
   Learning Repository. <https://doi.org/10.24432/C5DW2B>
