---
title: "Lesson 0 — Setup and the R basics you need"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
# This hidden chunk sets options for the whole document.
# echo = TRUE means "show the code in the final report".
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# What this lesson is for

Before building machine-learning (ML) models you need three things:

1. **The tools installed** (R, RStudio, and some add-on *packages*).
2. **How an R Markdown file works**, because every lesson is one.
3. **About 10 R ideas** you will see in every lesson. You do **not** need to
   become a programmer. You need to be able to *read* code and change it.

Time needed: about 1–2 hours.

# Install the software (one time only)

| Step | What | Where |
|------|------|-------|
| 1 | **R**, the language that runs the calculations (designed for R 4.6.1) | <https://cran.r-project.org> |
| 2 | **RStudio Desktop**, the program you type in | <https://posit.co/download/rstudio-desktop/> |
| 3 | **Packages**, free add-ons that contain the ML methods | Section 4 below |

> **Tip:** Install R first, then RStudio. RStudio finds R automatically.

# A tour of RStudio

When you open RStudio you see four panels:

| Panel | Usual position | What it does |
|-------|----------------|--------------|
| **Source / editor** | top-left | Where this `.Rmd` file is open. You write and edit here. |
| **Console** | bottom-left | Where code actually runs. Results and errors print here. |
| **Environment** | top-right | Lists the objects (datasets, models) currently in memory. |
| **Files / Plots / Help** | bottom-right | Your folders, your graphs, and the help pages. |

## Open the course as a *Project*

Double-click **`ML_in_R_course.Rproj`** in this folder. This opens RStudio
*inside* the course folder, so file paths like `data/heart_clean.rds` always
work. **Always open the course this way.**

## How an R Markdown (`.Rmd`) file works

An `.Rmd` file mixes three things:

1. **The header** (the part between the `---` lines at the top): title,
   author, output format.
2. **Normal text**, like this paragraph. You write your project's intent,
   methods and interpretation here, as in a paper.
3. **Code chunks**, the grey boxes that start with ```` ```{r} ````. Only
   these are run as R code.

Keyboard shortcuts you will use constantly:

| Action | Windows shortcut | Button |
|--------|------------------|--------|
| Run the current line or selection | `Ctrl + Enter` | "Run" |
| Run the whole chunk | `Ctrl + Shift + Enter` | green ▶ on the chunk |
| Run all chunks above this one | `Ctrl + Alt + P` | grey ▼ on the chunk |
| **Knit** (build the full HTML report) | `Ctrl + Shift + K` | "Knit" (ball of yarn) |
| Insert a new chunk | `Ctrl + Alt + I` | "+C" button |

> **How to learn with these files:** run each chunk **one at a time** with the
> green ▶, read the output, and only then move on. When you finish, click
> **Knit** to get a clean HTML report of everything.

# Install the packages (one time only)

A *package* is a bundle of functions written by other people. You install a
package **once** with `install.packages()`, and you *load* it **every
session** with `library()`. It's like installing an app on your phone once
and then opening it whenever you need it.

The chunk below has `eval=FALSE`, so it does **not** run when you Knit. Run
it **by hand once**: put the cursor inside the chunk and press
`Ctrl + Shift + Enter`. It can take 5–10 minutes.

```{r install-packages, eval=FALSE}
# c() means "combine". It builds a list of package names (text goes in quotes).
packages_needed <- c(
  "tidyverse",   # data handling (dplyr) and plots (ggplot2)
  "tidymodels",  # one consistent toolkit for ML: split, preprocess, tune, evaluate
  "gtsummary",   # publication-ready "Table 1" and regression tables
  "broom.helpers", # needed by gtsummary for regression tables
  "skimr",       # quick summary of every variable in a dataset
  "glmnet",      # penalized (LASSO / ridge) logistic regression
  "ranger",      # fast random forests
  "xgboost",     # XGBoost gradient boosting
  "lightgbm",    # LightGBM gradient boosting
  "bonsai"       # connects LightGBM to tidymodels
)

# Keep only the packages that are not installed yet.
# installed.packages() lists what you have; rownames() pulls out the names.
# setdiff(A, B) gives the items in A that are not in B.
missing_packages <- setdiff(packages_needed, rownames(installed.packages()))

# Install whatever is missing. If nothing is missing, this does nothing.
install.packages(missing_packages)
```

Check that everything loads. If a line gives an error like
`there is no package called 'xxx'`, run `install.packages("xxx")` in the
Console.

```{r load-check}
library(tidyverse)   # loads ggplot2, dplyr, readr, tidyr, and others
library(tidymodels)  # loads rsample, recipes, parsnip, tune, yardstick, and others
library(bonsai)      # makes LightGBM available inside tidymodels

# R.version.string is a built-in piece of text with your R version.
R.version.string
```

# The R ideas you need (and nothing more)

## Objects and the assignment arrow `<-`

`<-` means "store the result on the right under the name on the left". The
name then appears in the **Environment** panel.

```{r objects}
age <- 63            # store the number 63 under the name "age"
age                  # typing a name alone prints its value
age_in_months <- age * 12   # do maths with it and store the result
age_in_months
```

## Vectors: many values in one object

```{r vectors}
# c() combines values into a "vector" (a column of values of the same type)
cholesterol <- c(233, 286, 229, 250, 204)

mean(cholesterol)    # mean() is a function: name, then the input in brackets
length(cholesterol)  # how many values there are
cholesterol > 240    # compares each value; gives TRUE or FALSE for each one
```

## Functions and their *arguments*

A function is `name(argument1 = value, argument2 = value)`. To read the help
page for any function, type `?` and the name in the Console, e.g. `?mean`.

```{r functions}
# round() takes a number and how many digits to keep.
round(3.14159, digits = 2)

# If you give arguments in the documented order, you can drop the names:
round(3.14159, 2)
```

## Data frames (tibbles): your dataset

A **data frame** is a table: rows are patients (observations) and columns are
variables. The tidyverse version is called a **tibble**.

```{r dataframes}
# tibble() builds a small table. Each argument becomes a column.
patients <- tibble(
  id      = 1:4,                          # 1:4 means 1, 2, 3, 4
  age     = c(63, 67, 37, 41),
  sex     = c("Male", "Male", "Male", "Female"),
  disease = c("No", "Yes", "No", "No")
)

patients            # print the table
nrow(patients)      # number of rows
patients$age        # the $ sign pulls out one column
```

## Factors: categorical variables

A **factor** is R's way of storing a categorical variable (e.g. sex, disease
yes/no). ML functions need categorical outcomes stored as factors.

```{r factors}
patients$disease <- factor(patients$disease, levels = c("Yes", "No"))
levels(patients$disease)  # the categories, in their set order
```

> **Remember for later:** in **tidymodels**, the **first level** of the
> outcome factor is treated as the "event" (the thing we try to predict).
> That is why we put `"Yes"` first.

## The pipe `|>`: "and then..."

The pipe passes the result on its left into the function on its right. Read
it aloud as **"and then"**.

```{r pipe}
patients |>                      # take the patients table, AND THEN
  filter(age > 40) |>            # keep only rows where age is over 40, AND THEN
  select(id, age, disease)       # keep only these three columns
```

Main **dplyr** verbs (all used in the lessons):

| Verb | What it does |
|------|--------------|
| `filter()` | keep rows that meet a condition |
| `select()` | keep (or drop) columns |
| `mutate()` | create or change a column |
| `summarise()` | collapse many rows into a summary (mean, count...) |
| `group_by()` | do the next step separately for each group |
| `count()` | count rows per category |

## Plots with ggplot2

`ggplot()` builds a plot in layers joined with `+` (not with `|>`).

```{r ggplot-demo, fig.height=3, fig.width=5}
ggplot(patients, aes(x = disease, y = age)) +  # data, and which columns go on which axis
  geom_point(size = 3) +                        # draw points
  labs(title = "Age by disease status")         # add a title
```

## Reproducibility: `set.seed()`

ML uses randomness (random data splits, random trees). `set.seed(number)`
makes the "random" results **the same every time you run the code**, which
reproducible research requires.

```{r seed}
set.seed(123)
sample(1:10, 3)   # pick 3 random numbers between 1 and 10
set.seed(123)
sample(1:10, 3)   # same seed, so exactly the same 3 numbers
```

# When you get an error

1. **Read the last line of the red text.** It usually says what is wrong.
2. Common causes: a typo in a name, a missing bracket or comma, a package
   not loaded (run the `library()` chunk), or running chunks out of order.
3. Click **Session → Restart R**, then use the grey ▼ "run all chunks above".
4. Paste the error into a search engine or ask an AI assistant. Include the
   code that caused it.

# Check your understanding

1. What is the difference between `install.packages()` and `library()`?
2. What does `<-` do?
3. In `patients |> filter(age > 40)`, what does `|>` mean?
4. Why must we call `set.seed()` in ML research?

# Next

Open **`01_data_import_and_exploration.Rmd`**.
