---
title: "Lesson 1 — Getting real public data, cleaning it, and exploring it"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Project intent

> In a real study this section is your **introduction and aim**. Write it
> *before* touching the data. It stops you from "fishing" for results later.

**Research question.** Can routinely collected clinical and exercise-test
variables predict whether a patient has angiographically confirmed coronary
artery disease (CAD)?

**Why it matters.** Coronary angiography is invasive. A model that estimates
the probability of CAD from non-invasive data could help decide who needs it.

**Data source.** The *Cleveland Heart Disease* dataset from the UCI Machine
Learning Repository: 303 patients referred for angiography at the Cleveland
Clinic Foundation (Detrano et al., 1989). It is free and public:

* Janosi A, Steinbrunn W, Pfisterer M, Detrano R. *Heart Disease* [Dataset].
  UCI Machine Learning Repository, 1989. <https://doi.org/10.24432/C52P4X>

**Outcome.** CAD present (`Yes`) if any major vessel has more than 50%
diameter narrowing, otherwise `No`.

**What this lesson produces.** A clean dataset and a fixed **training/test
split**, saved to the `data/` folder and reused by every later lesson, so
that all models are compared on exactly the same patients.

# Load the packages

```{r packages}
library(tidyverse)   # data handling and plots
library(tidymodels)  # we only need rsample (the data splitting tool) from it today
library(skimr)       # skim(): a quick overview of every variable
library(gtsummary)   # tbl_summary(): publication-style "Table 1"
```

# Download the data from the web

The file is a plain text file with commas between values (a **CSV**), but it
has **no header row** (no column names) and uses `?` for missing values. We
tell R both things.

```{r download}
# The web address of the raw data file. We store it as text in an object.
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"

# Names for the 14 columns, in the order given by the dataset documentation.
column_names <- c("age", "sex", "chest_pain", "resting_bp", "cholesterol",
                  "fasting_bs", "rest_ecg", "max_hr", "exercise_angina",
                  "st_depression", "st_slope", "n_vessels", "thal", "num")

# read_csv() reads a CSV file (it can read directly from a web address).
heart_raw <- read_csv(
  data_url,
  col_names = column_names,   # use our names because the file has no header
  na = "?"                    # treat "?" as missing (NA in R)
)

# Save an untouched copy on your computer, so you never depend on the website.
# dir.create() makes a folder; showWarnings = FALSE stops a warning if it exists.
dir.create("data", showWarnings = FALSE)
write_csv(heart_raw, "data/heart_raw.csv")

# Look at the first rows. head() shows the first 6 by default.
head(heart_raw)
```

> **Good research practice:** keep the raw data **unchanged**. Every cleaning
> step happens in code, so anyone can repeat it exactly.

# Understand the variables (the "data dictionary")

Every value is stored as a number, but many are really **categories**. The
dataset documentation tells us what each code means:

| Column | Meaning | Coding |
|--------|---------|--------|
| `age` | Age (years) | number |
| `sex` | Sex | 0 = female, 1 = male |
| `chest_pain` | Chest pain type | 1 = typical angina, 2 = atypical angina, 3 = non-anginal, 4 = asymptomatic |
| `resting_bp` | Resting systolic blood pressure (mmHg) | number |
| `cholesterol` | Serum cholesterol (mg/dL) | number |
| `fasting_bs` | Fasting blood sugar > 120 mg/dL | 0 = no, 1 = yes |
| `rest_ecg` | Resting ECG | 0 = normal, 1 = ST-T abnormality, 2 = left ventricular hypertrophy |
| `max_hr` | Maximum heart rate on exercise test | number |
| `exercise_angina` | Exercise-induced angina | 0 = no, 1 = yes |
| `st_depression` | ST depression on exercise relative to rest (mm) | number |
| `st_slope` | Slope of the peak exercise ST segment | 1 = up, 2 = flat, 3 = down |
| `n_vessels` | Major vessels coloured by fluoroscopy | 0–3 |
| `thal` | Thallium stress test | 3 = normal, 6 = fixed defect, 7 = reversible defect |
| `num` | Angiographic diagnosis | 0 = no CAD, 1–4 = CAD |

# Clean the data

Here we turn number codes into labelled **factors**, create the outcome, and
remove the original `num` column. Leaving `num` in would let the model "cheat",
because it *is* the answer. This mistake is called **target leakage**.

```{r clean}
heart <- heart_raw |>                 # start from the raw data, AND THEN
  mutate(                             # create or overwrite columns:

    # factor(x, levels = codes, labels = names) turns codes into categories
    sex = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),

    chest_pain = factor(chest_pain, levels = 1:4,
                        labels = c("Typical angina", "Atypical angina",
                                   "Non-anginal pain", "Asymptomatic")),

    fasting_bs = factor(fasting_bs, levels = c(0, 1), labels = c("No", "Yes")),

    # rest_ecg code 1 has only 4 patients. Groups that small make unstable
    # estimates, so we merge codes 1 and 2 into "Abnormal".
    # if_else(condition, value if TRUE, value if FALSE)
    rest_ecg = factor(if_else(rest_ecg == 0, "Normal", "Abnormal"),
                      levels = c("Normal", "Abnormal")),

    exercise_angina = factor(exercise_angina, levels = c(0, 1), labels = c("No", "Yes")),

    st_slope = factor(st_slope, levels = 1:3,
                      labels = c("Upsloping", "Flat", "Downsloping")),

    thal = factor(thal, levels = c(3, 6, 7),
                  labels = c("Normal", "Fixed defect", "Reversible defect")),

    # The OUTCOME: "Yes" if num is 1 to 4, "No" if num is 0.
    # "Yes" is put FIRST so tidymodels treats it as the event of interest.
    disease = factor(if_else(num > 0, "Yes", "No"), levels = c("Yes", "No"))
  ) |>
  select(-num)                        # the minus sign drops the num column (avoids leakage)

glimpse(heart)   # one line per column: its name, type and first values
```

Why did we keep a small category in mind? Because we checked the counts.
`count()` is how you would find this yourself:

```{r check-counts}
count(heart_raw, rest_ecg)   # how many patients per original rest_ecg code
```

# Explore the data (EDA)

**Exploratory data analysis (EDA)** means looking at the data before
modelling. Look at missing values, distributions, odd values, and the outcome
balance.

## Quick overview

```{r skim}
skim(heart)   # for each variable: type, number missing, mean, sd, percentiles, histogram
```

## Missing values

```{r missing}
# is.na() is TRUE where a value is missing. colSums() adds up the TRUEs per column.
colSums(is.na(heart))
```

Only 6 values are missing (`n_vessels` 4, `thal` 2). We will **not** delete
these patients. Instead the models will **impute** (fill in) these values
*using only training data* (explained in Lesson 2).

## Is the outcome balanced?

```{r outcome-balance, fig.height=3, fig.width=5}
heart |>
  count(disease) |>                          # number of patients per outcome
  mutate(percent = round(100 * n / sum(n), 1))  # turn counts into percentages

ggplot(heart, aes(x = disease, fill = disease)) +
  geom_bar() +                                # bar height = count of patients
  labs(x = "Coronary artery disease", y = "Number of patients") +
  theme_minimal() +
  theme(legend.position = "none")             # the legend would repeat the x-axis
```

About 46% have CAD, so the outcome is fairly **balanced**. With a rare outcome
(say 5%), accuracy becomes misleading and you would need extra care (see the
glossary in `README.md`).

## Table 1: patient characteristics by outcome

`tbl_summary()` makes the classic "Table 1" of a clinical paper in one line.

```{r table1}
heart |>
  tbl_summary(
    by = disease,                         # one column per outcome group
    missing_text = "Missing"              # how missing values are labelled
  ) |>
  add_overall() |>                        # add an "Overall" column
  add_p() |>                              # add p-values (group comparison tests)
  bold_labels()                           # make variable names bold
```

## Distributions of the numeric variables

To plot all numeric variables together, we first reshape the data from
**wide** (one column per variable) to **long** (one row per patient-variable
pair). `pivot_longer()` does this.

```{r numeric-plots, fig.height=5, fig.width=9}
heart |>
  select(where(is.numeric), disease) |>   # keep numeric columns plus the outcome
  pivot_longer(
    cols = -disease,                      # reshape every column except disease
    names_to = "variable",                # old column names go in "variable"
    values_to = "value"                   # their values go in "value"
  ) |>
  ggplot(aes(x = value, fill = disease)) +
  geom_density(alpha = 0.5) +             # smoothed histogram; alpha = transparency
  facet_wrap(~ variable, scales = "free") + # one small panel per variable
  theme_minimal()
```

**What to look for:** variables whose curves are clearly separated (for
example `max_hr`, `st_depression` and `n_vessels`) are likely strong
predictors.

## Categorical variables vs the outcome

```{r categorical-plots, fig.height=6, fig.width=9}
heart |>
  select(where(is.factor)) |>             # keep only factor (categorical) columns
  pivot_longer(cols = -disease, names_to = "variable", values_to = "category") |>
  ggplot(aes(x = category, fill = disease)) +
  geom_bar(position = "fill") +           # "fill" = bars show proportions, not counts
  facet_wrap(~ variable, scales = "free_x") +
  labs(y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))  # tilt the labels
```

# The most important step in ML: split the data

## Why split?

A model that is tested on the same patients it learned from will look much
better than it really is, because it partly **memorises** them. This is
called **overfitting**. To get an honest estimate, we hide some patients from
the model and test on them at the very end.

* **Training set (75%)**: used to build and tune the models.
* **Test set (25%)**: locked away and used **once**, at the very end.

`strata = disease` keeps the same proportion of CAD in both sets
(**stratified split**).

```{r split}
set.seed(2024)   # makes the random split identical every time

heart_split <- initial_split(
  heart,               # the data to split
  prop = 0.75,         # 75% training, 25% test
  strata = disease     # keep the outcome proportion similar in both parts
)

heart_split          # shows <Training/Testing/Total> numbers

heart_train <- training(heart_split)   # pull out the training patients
heart_test  <- testing(heart_split)    # pull out the test patients

# Check that the outcome proportion is similar in both sets
bind_rows(
  train = count(heart_train, disease),
  test  = count(heart_test, disease),
  .id = "set"                          # adds a column saying which set each row came from
) |>
  group_by(set) |>
  mutate(percent = round(100 * n / sum(n), 1))
```

# Save the results for the next lessons

`.rds` is R's own file format. It keeps factor levels and the split object
exactly as they are.

```{r save}
saveRDS(heart, "data/heart_clean.rds")        # the clean full dataset
saveRDS(heart_split, "data/heart_split.rds")  # the split (which patients are train/test)
```

# Key takeaways

* Write the **research question and outcome definition before** looking at
  results.
* Keep raw data unchanged. Do all cleaning **in code**.
* Remove anything that "contains the answer" (**target leakage**).
* **Split before modelling.** The test set is touched once, at the end.

# Check your understanding

1. Why did we remove the `num` column?
2. What would happen to our accuracy estimate if we tested on training data?
3. What does `strata = disease` do?

# Try it yourself

1. Change `prop = 0.75` to `0.8` and rerun the split. How many test patients
   are there now? (Change it back afterwards.)
2. Add `cholesterol` vs `age` as a scatter plot coloured by `disease`. Hint:
   `ggplot(heart, aes(x = age, y = cholesterol, colour = disease)) + geom_point()`.

# Next

Open **`02_logistic_regression.Rmd`**.
