---
title: "Lesson 5 — Gradient boosting with LightGBM"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The concept

**LightGBM** (Light Gradient Boosting Machine; Ke et al., 2017, Microsoft) is
gradient boosting, the same idea as Lesson 4, with some engineering
differences:

| | XGBoost (default) | LightGBM |
|---|---|---|
| How trees grow | **Level-wise**: splits every node at one depth before going deeper | **Leaf-wise**: always splits the single leaf that reduces error most |
| Speed on big data | Fast | Usually **faster**: bins values into histograms |
| Categorical variables | Need dummy columns | Handled **natively** |
| Missing values | Handled natively | Handled natively |
| Main complexity setting | `tree_depth` | `num_leaves` (maximum leaves per tree) |

```
Level-wise (XGBoost)            Leaf-wise (LightGBM)
        o                              o
      /   \                          /   \
     o     o                        o     o
    / \   / \                            / \
   o   o o   o                          o   o
                                           / \
                                          o   o
```

**When does LightGBM shine?** Large datasets (tens of thousands of rows or
more) with many variables, such as registry data, EHR extracts or omics. On a
300-patient dataset like ours, its speed advantage is invisible and leaf-wise
growth can overfit, so keep trees small.

## Key hyperparameters

| tidymodels name | LightGBM name | Meaning |
|---|---|---|
| `trees` | `num_iterations` | Number of boosting rounds |
| `learn_rate` | `learning_rate` | Step size of each correction |
| `min_n` | `min_data_in_leaf` | Minimum patients per leaf. **Important for small data** |
| `num_leaves` (engine argument) | `num_leaves` | Maximum leaves per tree. Controls complexity |

# Load packages and data

```{r packages}
library(tidyverse)
library(tidymodels)
library(bonsai)       # adds the "lightgbm" engine to tidymodels
tidymodels_prefer()

heart_split <- readRDS("data/heart_split.rds")
heart_train <- training(heart_split)
heart_test  <- testing(heart_split)

set.seed(2024)
heart_folds <- vfold_cv(heart_train, v = 10, strata = disease)

heart_metrics <- metric_set(roc_auc, brier_class, accuracy, sensitivity, specificity)
```

# The recipe: almost nothing to do

LightGBM handles **categories and missing values natively**, so the recipe
only declares the outcome and predictors. No steps are needed.

```{r recipe}
lgb_recipe <- recipe(disease ~ ., data = heart_train)   # no preprocessing steps
```

# Model specification with tuning

```{r lgb-spec}
lgb_spec <- boost_tree(
  trees      = tune(),
  learn_rate = tune(),
  min_n      = tune()
) |>
  set_engine(
    "lightgbm",
    num_leaves  = tune(),   # an engine-specific setting we also tune
    num_threads = 1,        # use 1 CPU core, so results are exactly reproducible
    verbose     = -1        # hide LightGBM's progress messages
  ) |>
  set_mode("classification")

lgb_wf <- workflow() |>
  add_recipe(lgb_recipe) |>
  add_model(lgb_spec)
```

## Search ranges and grid

```{r lgb-params}
lgb_params <- lgb_wf |>
  extract_parameter_set_dials() |>
  update(
    trees      = trees(range = c(100, 1000)),
    learn_rate = learn_rate(range = c(-3, -0.5)),   # 0.001 to 0.32
    min_n      = min_n(range = c(5, 40)),           # at least 5 patients per leaf
    num_leaves = num_leaves(range = c(4, 31))       # small trees for small data
  )

set.seed(2024)
lgb_grid <- grid_space_filling(lgb_params, size = 30)
```

```{r lgb-tune}
lgb_tune <- tune_grid(
  lgb_wf,
  resamples = heart_folds,
  grid      = lgb_grid,
  metrics   = heart_metrics
)
```

```{r lgb-tune-plot, fig.height=6, fig.width=10}
autoplot(lgb_tune, metric = "roc_auc")
```

```{r lgb-best}
show_best(lgb_tune, metric = "roc_auc", n = 5)
best_lgb <- select_best(lgb_tune, metric = "roc_auc")
best_lgb
```

# Final model and test-set evaluation

```{r lgb-final}
lgb_final_wf <- finalize_workflow(lgb_wf, best_lgb)

lgb_final <- last_fit(lgb_final_wf, split = heart_split, metrics = heart_metrics)

collect_metrics(lgb_final)
```

```{r lgb-roc, fig.height=4, fig.width=4}
collect_predictions(lgb_final) |>
  roc_curve(truth = disease, .pred_Yes) |>
  autoplot()
```

# Variable importance

Because LightGBM used the categories directly, importance is reported per
**original variable** (e.g. `thal`), not per dummy column (e.g.
`thal_Reversible.defect`).

```{r lgb-importance, fig.height=4, fig.width=7}
lgb_booster <- extract_fit_engine(lgb_final)          # the raw LightGBM model

lgb_importance <- lightgbm::lgb.importance(lgb_booster)

ggplot(lgb_importance, aes(x = Gain, y = reorder(Feature, Gain))) +
  geom_col(fill = "darkorange") +
  labs(x = "Importance (gain)", y = NULL, title = "LightGBM variable importance") +
  theme_minimal()
```

# Save results for Lesson 6

```{r save}
collect_predictions(lgb_final) |>
  mutate(model = "LightGBM") |>
  select(model, .row, disease, .pred_Yes, .pred_class) |>  # .row = patient ID
  saveRDS("results/pred_lgb.rds")

collect_metrics(lgb_tune) |>
  filter(.config == best_lgb$.config) |>
  mutate(model = "LightGBM") |>
  select(model, .metric, mean, std_err) |>
  saveRDS("results/cv_lgb.rds")
```

# Key takeaways

* LightGBM is gradient boosting with **leaf-wise** growth and **histogram**
  binning. It is very fast on large data.
* It handles **categorical and missing** values without preprocessing.
* On small data, control complexity with `num_leaves` and `min_n` to avoid
  overfitting.
* Thanks to tidymodels, switching algorithm meant changing only the **model
  specification** (and simplifying the recipe). The rest of the code is
  identical.

# Check your understanding

1. What is the difference between level-wise and leaf-wise tree growth?
2. Why do we not need `step_dummy()` for LightGBM?
3. When would you choose LightGBM over XGBoost?

# Try it yourself

1. Change the `num_leaves` range to `c(2, 8)`. Does test AUC improve or get
   worse?
2. Add `tree_depth = tune()` to `boost_tree()` and a range
   `tree_depth = tree_depth(range = c(2, 6))` in `update()`.

# Next

Open **`06_model_comparison.Rmd`**.
