---
title: "Lesson 6 — Comparing models and reporting them like a researcher"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Goal

You have trained five models on **the same training data**, tuned them on
**the same CV folds**, and tested them on **the same 76 test patients**.
Now we compare them properly. That means looking at more than "which AUC is
highest":

1. **Discrimination**: can the model rank patients by risk? (AUC, with 95% CI)
2. **Calibration**: are the predicted probabilities *correct*? If the model
   says 30%, do about 30% of those patients have CAD?
3. **Clinical usefulness**: would using the model lead to better decisions
   than "treat everyone" or "treat no one"? (decision curve analysis)
4. **Uncertainty**: are the differences real or just noise?

> Run Lessons 2–5 first. They save their results in the `results/` folder.

# Load packages and results

```{r packages}
library(tidyverse)
library(tidymodels)
tidymodels_prefer()

# list.files() finds files whose names match a pattern; full.names gives the full path
pred_files <- list.files("results", pattern = "^pred_", full.names = TRUE)
cv_files   <- list.files("results", pattern = "^cv_",   full.names = TRUE)

# map() applies readRDS to each file; list_rbind() stacks the results into one table
all_preds <- pred_files |> map(readRDS) |> list_rbind()
all_cv    <- cv_files   |> map(readRDS) |> list_rbind()

# Fix the order models appear in tables and plots (simplest to most complex)
model_order <- c("Logistic regression", "LASSO logistic", "Random forest", "XGBoost", "LightGBM")
all_preds <- all_preds |> mutate(model = factor(model, levels = model_order))
all_cv    <- all_cv    |> mutate(model = factor(model, levels = model_order))

count(all_preds, model)   # each model should have 76 test predictions
```

# Cross-validation performance (training data)

```{r cv-table}
all_cv |>
  arrange(model) |>                                          # sort rows in model_order
  mutate(value = sprintf("%.3f (%.3f)", mean, std_err)) |>  # format as "mean (SE)"
  select(model, .metric, value) |>
  pivot_wider(names_from = .metric, values_from = value) |>  # one column per metric
  knitr::kable(caption = "10-fold cross-validation on the training set: mean (standard error)")
```

# Test-set performance with 95% confidence intervals

The test set has only 76 patients, so every estimate is **uncertain**. We
quantify this with the **bootstrap**:

1. Draw 76 patients **with replacement** from the test set (some appear twice,
   some not at all).
2. Compute every model's AUC and Brier score on this resample.
3. Repeat 1,000 times. The middle 95% of the results is the 95% CI.

Because every model is evaluated on the **same** resampled patients each time
(a *paired* bootstrap), we can also get a CI for the **difference** between
two models.

```{r bootstrap}
set.seed(2024)
patient_ids <- unique(all_preds$.row)   # the 76 test-patient IDs

# function(i) { ... } defines a small function. map() runs it for i = 1, 2, ..., 1000.
boot_results <- map(1:1000, function(i) {
  resampled <- tibble(.row = sample(patient_ids, replace = TRUE))  # step 1: resample patients
  resampled |>
    left_join(all_preds, by = ".row", relationship = "many-to-many") |>  # attach each model's predictions
    group_by(model) |>
    summarise(                                          # step 2: metrics per model
      auc   = roc_auc_vec(disease, .pred_Yes),
      brier = brier_class_vec(disease, .pred_Yes),
      .groups = "drop"
    ) |>
    mutate(boot = i)                                    # remember which repetition this was
}) |>
  list_rbind()                                          # stack the 1,000 results

# Point estimates on the real (not resampled) test set
point_estimates <- all_preds |>
  group_by(model) |>
  summarise(auc   = roc_auc_vec(disease, .pred_Yes),
            brier = brier_class_vec(disease, .pred_Yes),
            sens  = sensitivity_vec(disease, .pred_class),
            spec  = specificity_vec(disease, .pred_class))

# Step 3: 95% CIs = 2.5th and 97.5th percentiles of the bootstrap results
ci <- boot_results |>
  group_by(model) |>
  summarise(auc_low   = quantile(auc, 0.025),   auc_high   = quantile(auc, 0.975),
            brier_low = quantile(brier, 0.025), brier_high = quantile(brier, 0.975))

test_table <- left_join(point_estimates, ci, by = "model")

test_table |>
  transmute(                                    # transmute = mutate + keep only the new columns
    Model         = model,
    `AUC (95% CI)`   = sprintf("%.3f (%.3f–%.3f)", auc, auc_low, auc_high),
    `Brier (95% CI)` = sprintf("%.3f (%.3f–%.3f)", brier, brier_low, brier_high),
    Sensitivity   = sprintf("%.2f", sens),
    Specificity   = sprintf("%.2f", spec)
  ) |>
  knitr::kable(caption = "Performance on the held-out test set (n = 76)")
```

## Is any model *really* better than logistic regression?

```{r paired-difference}
boot_results |>
  select(boot, model, auc) |>
  pivot_wider(names_from = model, values_from = auc) |>   # one column per model
  mutate(across(                                           # for each model column...
    -c(boot, `Logistic regression`),
    ~ .x - `Logistic regression`                           # ...subtract the logistic AUC
  )) |>
  select(-boot, -`Logistic regression`) |>
  pivot_longer(everything(), names_to = "model", values_to = "auc_difference") |>
  group_by(model) |>
  summarise(
    difference = median(auc_difference),
    ci_low     = quantile(auc_difference, 0.025),
    ci_high    = quantile(auc_difference, 0.975)
  ) |>
  knitr::kable(digits = 3, caption = "AUC difference vs logistic regression (paired bootstrap)")
```

**How to read this:** if the 95% CI of the difference includes 0, the test
set gives no evidence that the ML model discriminates better than logistic
regression. With small samples like this one, that is the usual result, and
it matches a well-known systematic review: ML did not outperform logistic
regression for clinical prediction on average (Christodoulou et al., *J Clin
Epidemiol* 2019).

# ROC curves

```{r roc-all, fig.height=5, fig.width=6}
all_preds |>
  group_by(model) |>                       # one curve per model
  roc_curve(truth = disease, .pred_Yes) |>
  autoplot() +
  labs(title = "ROC curves on the test set")
```

# Calibration: are the probabilities right?

A model can rank patients well (high AUC) but still give **wrong
probabilities**, for example predicting 80% for patients whose real risk is 50%.
Clinicians act on probabilities, so calibration matters.

**Calibration plot:** sort patients into groups by predicted risk, then
compare the **average predicted** risk to the **observed** proportion with CAD
in each group. A perfect model lies on the diagonal.

```{r calibration-plot, fig.height=6, fig.width=9}
calibration_data <- all_preds |>
  group_by(model) |>
  mutate(risk_group = ntile(.pred_Yes, 5)) |>   # 5 groups of ~15 patients, from lowest to highest predicted risk
  group_by(model, risk_group) |>
  summarise(
    predicted = mean(.pred_Yes),                # average predicted probability in the group
    observed  = mean(disease == "Yes"),         # actual proportion with CAD in the group
    n         = n(),
    .groups = "drop"
  )

ggplot(calibration_data, aes(x = predicted, y = observed)) +
  geom_abline(linetype = "dashed", colour = "grey50") +   # the perfect-calibration line
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue", size = 2) +
  facet_wrap(~ model) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +           # same scale on both axes
  labs(x = "Predicted probability", y = "Observed proportion with CAD",
       title = "Calibration on the test set (5 risk groups)") +
  theme_minimal()
```

## Calibration intercept and slope

* **Calibration slope** = 1 is ideal. **< 1** means predictions are too
  extreme (typical of overfitting). **> 1** means too modest.
* **Calibration intercept** (calibration-in-the-large) = 0 is ideal. It shows
  whether risks are systematically too high or too low.

```{r calibration-stats}
calibration_stats <- function(truth, prob) {
  y  <- as.integer(truth == "Yes")                       # 1 = CAD, 0 = no CAD
  lp <- qlogis(pmin(pmax(prob, 0.001), 0.999))           # probability to log-odds (avoid 0 and 1)
  slope     <- coef(glm(y ~ lp, family = binomial))[2]   # logistic regression of outcome on log-odds
  intercept <- coef(glm(y ~ offset(lp), family = binomial))[1]  # intercept with slope fixed at 1
  tibble(cal_intercept = intercept, cal_slope = slope)
}

all_preds |>
  group_by(model) |>
  summarise(calibration_stats(disease, .pred_Yes)) |>
  knitr::kable(digits = 2, caption = "Calibration intercept (ideal 0) and slope (ideal 1)")
```

# Clinical usefulness: decision curve analysis (DCA)

Suppose we refer a patient for angiography if their predicted risk is above
a **threshold** `t`. Choosing `t` reflects how we weigh a missed case against
an unnecessary angiogram. **Net benefit** (Vickers & Elkin, 2006) counts true
positives and subtracts false positives, weighted by that threshold:

`net benefit = TP/n − FP/n × t / (1 − t)`

A model is useful at a threshold if its net benefit is above both reference
strategies: **"refer everyone"** and **"refer no one"** (net benefit = 0).

```{r dca, fig.height=5, fig.width=8}
thresholds <- tibble(threshold = seq(0.05, 0.80, by = 0.01))
prevalence <- mean(all_preds$disease[all_preds$model == "Logistic regression"] == "Yes")

dca <- all_preds |>
  cross_join(thresholds) |>                 # every prediction paired with every threshold
  group_by(model, threshold) |>
  summarise(
    tp_rate = sum(.pred_Yes >= threshold & disease == "Yes") / n(),  # true positives / n
    fp_rate = sum(.pred_Yes >= threshold & disease == "No")  / n(),  # false positives / n
    .groups = "drop"
  ) |>
  mutate(
    net_benefit = tp_rate - fp_rate * threshold / (1 - threshold),   # the net-benefit formula
    model = as.character(model)
  )

treat_all <- thresholds |>
  mutate(model = "Refer everyone",
         net_benefit = prevalence - (1 - prevalence) * threshold / (1 - threshold))

bind_rows(dca, treat_all) |>
  ggplot(aes(x = threshold, y = net_benefit, colour = model)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, colour = "black") +           # "refer no one"
  coord_cartesian(ylim = c(-0.05, prevalence + 0.05)) +    # zoom to the useful range
  labs(x = "Risk threshold for referral", y = "Net benefit", colour = NULL,
       title = "Decision curve analysis (test set)") +
  theme_minimal()
```

# Choosing a final model

There is no automatic answer. Weigh:

* **Performance**, *including uncertainty*. Overlapping CIs mean "no
  demonstrated difference".
* **Calibration**: a well-calibrated simpler model often beats a sharper but
  mis-calibrated complex one.
* **Interpretability and implementation**: logistic regression can be written
  as a score on paper. A boosted model needs software.
* **Parsimony**: when performance is similar, prefer the simpler model.

Note also that cross-validated AUCs (about 0.91–0.93) are higher than test AUCs. With
76 test patients the test estimate is noisy, and we selected the best tuning
values *on* those CV results (a small optimistic bias). This is one reason why
**external validation** in a different hospital or time period is the gold
standard.

# Reporting: write results directly from the code

R Markdown can put numbers from your code **inside your text** with inline
code: a backtick, the letter r, a space, your code, and a closing backtick.
The numbers update automatically if the data change, so there are no
copy-paste errors.

```{r best-model}
best <- test_table |> slice_max(auc, n = 1)   # the row with the highest test AUC
```

> **Example results sentence (rendered from code):** On the held-out test set
> (n = `r nrow(filter(all_preds, model == "Logistic regression"))`), the
> highest AUC was obtained by `r best$model` (`r sprintf("%.2f", best$auc)`;
> 95% CI `r sprintf("%.2f", best$auc_low)`–`r sprintf("%.2f", best$auc_high)`).
> Logistic regression achieved an AUC of
> `r sprintf("%.2f", test_table$auc[test_table$model == "Logistic regression"])`.

## Reporting checklist: TRIPOD+AI

For prediction-model papers, follow **TRIPOD+AI** (Collins et al., *BMJ*
2024; <https://www.tripod-statement.org>). The key items, and where they are
in this course:

| TRIPOD+AI item | Where in the course |
|---|---|
| Data source, eligibility, dates | Lesson 1 *Project intent* |
| Outcome definition | Lesson 1 *Clean the data* |
| Predictors and how they were handled | Lesson 1 data dictionary, recipes |
| Missing data handling | Recipes (imputation within resampling) |
| Sample size / number of events | Lesson 1 outcome balance, Lesson 2 caution |
| Model type, hyperparameter tuning | Lessons 2–5 (grids, 10-fold CV) |
| Performance: discrimination **and** calibration | This lesson |
| Clinical utility | Decision curve analysis (this lesson) |
| Uncertainty (CIs) | Bootstrap (this lesson) |
| Code and software availability | `sessionInfo()` below + share your `.Rmd` |

# Reproducibility information

```{r session-info}
sessionInfo()   # prints R version and package versions. Always include this in research
```

# Key takeaways

* Compare models on **the same patients**, with **CIs**, not just point
  estimates.
* Report **discrimination, calibration and clinical utility**.
* On small tabular datasets, complex ML often **does not beat** logistic
  regression. Report that honestly.
* Use inline R code so your text and results never disagree.

# Next

Open **`07_research_project_TEMPLATE.Rmd`**, a complete, reusable template
that applies everything to a **new** public dataset. Copy it for your own
studies.
